---
title: "Reproducing the C-sequestration case study with rCTOOL"
subtitle: "A worked implementation of the manuscript's AGWP/GWP accounting framework, on top of the rCTOOL soil-carbon model"
author: "Draft computational appendix (contact: Franca Giannini Kurina)"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
                       fig.width = 7.5, fig.height = 5, dpi = 150)
```

## What this document does

This is a fully reproducible implementation of the **Case study** section of
the manuscript *"C sequestration accounting"*, built on the
[**rCTOOL**](https://github.com/francagiannini/rCTOOL) package (the R
implementation of the C-TOOL soil carbon model, Taghizadeh-Toosi et al.,
2014). It reproduces:

* **Fig. 6** - CO2 emissions in the 10 years following a single 1000 kg C
  input (plant material at two C:N ratios, manure, and plant material under
  a cooler climate)
* **Fig. 7** - the associated N2O(-N) emissions
* **Fig. 8** - instantaneous AGWP-scale warming from the sequestered-C CO2
  emissions vs. an immediate (pulse) emission of the same mass
* **Fig. 9** - the same, for the N2O emitted from mineralisation of the
  co-input organic N
* **Table 2** - the net GWP effect (C sequestration offset + N2O emission)
  over a 100-year time horizon, in kg CO2e
* a **bonus scenario** (from the Discussion, the unfinished "Figure xx")
  comparing the conventional IPCC stock-change method against the
  AGWP-based method for a *repeated* annual C input

rCTOOL supplies the carbon-turnover engine only (the FOM/HUM/ROM pool
kinetics). Three things the manuscript adds on top of plain C-TOOL are **not**
part of the rCTOOL package and are implemented in `R/ghg_functions.R`,
sourced below:

1. a **single-layer** wrapper around rCTOOL's topsoil pool functions
   (the paper's Methods: transport parameter set to zero, and the HUM
   fraction rCTOOL would otherwise ship to the subsoil recycled back into
   topsoil HUM);
2. a **nitrogen-mineralisation / N2O** sub-model (Eq. 6-8) built on the
   carbon fluxes rCTOOL computes;
3. the **AGWP/GWP metric framework** itself (Eq. 1-9), after Petersen et
   al. (2013) and Aamaas et al. (2013), using the CO2 impulse-response
   function of Joos et al. (2013) and the N2O perturbation lifetime of
   Forster et al. (2021).

```{r libraries}
# install.packages("remotes")
# remotes::install_github("francagiannini/rCTOOL")
# install.packages(c("dplyr", "ggplot2", "tidyr", "knitr"))

library(rCTOOL)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)

source("R/ghg_functions.R")
theme_set(theme_classic(base_size = 12))
```

## A note on Table 1 of the draft

The draft's Table 1 lists $k_{HUM} = 0.028$ kg (kg C)^-1^ (confirmed by
inspecting the manuscript's own .docx table cells directly, so this is not a
transcription artefact on our side). Taken literally, that value decomposes
the HUM pool about 10x faster than rCTOOL's own published default
($k_{hum} = 0.0028$; see the package's `basic_example`/README), and it
produces C-sequestration GWP offsets 3-4x smaller in magnitude than the ones
quoted in the manuscript text.

Using $k_{HUM} = 0.0028$ instead - and measuring "years after input" from
$t=0$ at the *start* of the input year - reproduces the manuscript's quoted
C-sequestration offsets to within about 1% for three of the four scenarios,
and the N2O effective GWPs to within about 3-10% (validation numbers below,
section "Sanity checks"). That is a strong enough match to suggest "0.028"
in the draft is missing a leading zero. `k_hum` is set from a single named
constant so this is a one-line change if the table turns out to be correct
after all.

**A note on the installed rCTOOL version:** this was developed against
rCTOOL's GitHub "main" branch (v3.0.0); the CRAN release (v3.1.0) renamed
the seasonal-temperature argument of `FOM_top_calculations()` /
`HUM_top_calculations()` / `ROM_top_calculations()` from `t_range` to
`amplitude`, and rewrote the internal soil-temperature function to add a
real damped seasonal sine term rather than the GitHub version's
near-inert one. `run_ctool_pulse()` targets the CRAN argument name and,
as before, passes `amplitude = 0` deliberately (soil temperature = air
temperature, no depth lag): using the package's own auto-estimated
amplitude instead moves the C-sequestration offsets *further* from the
manuscript's quoted numbers (by a further 6-17%, worst for the
cool-climate scenario), so `amplitude = 0` remains the better-supported
choice. Because `amplitude = 0` zeroes out the damping term regardless of
thermal diffusivity, this choice is unaffected by the CRAN package's
default `temp_th_diff` differing 10x from the Table 1 value.

```{r table1-params}
k_hum_value <- 0.0028   # see note above; set to 0.028 for the draft's literal value

s_config <- soil_config(
  Csoil_init = 0,  Cproptop = 1,               # unused by the pulse wrapper (pools start at 0)
  f_hum_top = 0.4, f_rom_top = 0.4,             # unused
  f_hum_sub = 0.4, f_rom_sub = 0.4,             # unused (no subsoil in the single-layer wrapper)
  clay_top = 0.10, clay_sub = 0.10,             # Table 1: X = 0.10 g/g
  phi = 0.035,             # legacy parameter, unused by CRAN rCTOOL's soil-temperature function
  temp_th_diff = 0.035e-6, # Table 1: "0.035 x10^-6" - inert here since amplitude = 0 (see note above)
  f_co2  = 0.628,      # Table 1
  f_romi = 0.012,      # Table 1 (f_ROM)
  k_fom  = 0.12,       # Table 1
  k_hum  = k_hum_value,
  k_rom  = 3.85e-5,    # Table 1: "3.85 x10^-5"
  ftr    = 0           # single-layer configuration: transport parameter tF = 0
)
s_config$EF_N2O <- 0.01   # Table 1: kg N2O-N (kg N mineralised)-1

CN_hum <- 11.0; CN_rom <- 11.0   # Table 1

T_standard <- c(19, 22, 17, 16, 11, 11, 9, 8, 10, 13, 17, 19)
T_cool     <- c(14.8, 16.5, 16.0, 12.8, 9.0, 4.0, 1.8, 0.89, 1.17, 3.5, 6.0, 10.5)
stopifnot(abs(mean(T_standard) - 14.33) < 0.01, abs(mean(T_cool) - 8.08) < 0.01)
```

The monthly temperature *means* (14.33 and 8.08 &deg;C) match the 14.3/8.1
&deg;C quoted in the text for the standard and "cool climate" scenarios,
confirming the table's values were transcribed correctly.

### `f_man_humification`

This parameter (the fraction of manure C humified immediately on input) is
not given a numeric value anywhere in the draft. It is cross-checked instead
against a fact stated in the text: with the manuscript's parametrisation,
the manure's FOM sub-pool ends up with a C:N ratio of "17.9:1" (not the bulk
manure ratio of 16:1). Solving for the value of `f_man_humification` that
reproduces that number:

```{r fman-check}
f_man_humification <- 0.192   # also rCTOOL's own real-world (Askov) example value

Cman <- 1000; CNman <- 16
N_tot <- Cman / CNman
N_hum <- f_man_humification * Cman / CN_hum
CN_fom_check <- (Cman * (1 - f_man_humification)) / (N_tot - N_hum)
cat(sprintf("Implied manure FOM C:N ratio at f_man_humification = %.3f: %.1f (text says 17.9)\n",
            f_man_humification, CN_fom_check))
```

`f_man_humification = 0.192` reproduces the text's 17.9:1 almost exactly,
and is also rCTOOL's own default from its real-world Askov straw-incorporation
example, so it is used throughout below.

## Running the four case-study scenarios

```{r run-scenarios}
H <- 100            # time horizon, years (Eq. 1)
n_years_sim <- 150  # simulate beyond H so nothing past the horizon is missed
Cin_total <- 1000    # kg C, the manuscript's single-input case study

scenario_defs <- list(
  "Plant (C:N 16:1)"    = list(CN = 16, is_manure = FALSE, temps = T_standard),
  "Plant (C:N 50:1)"    = list(CN = 50, is_manure = FALSE, temps = T_standard),
  "Manure"               = list(CN = 16, is_manure = TRUE,  temps = T_standard),
  "Plant, cool climate"  = list(CN = 16, is_manure = FALSE, temps = T_cool)
)

sims <- lapply(scenario_defs, function(sc) {
  run_ctool_pulse(
    Cin_total = Cin_total, is_manure = sc$is_manure,
    f_man_humification = f_man_humification,
    CN_input = sc$CN, CN_hum = CN_hum, CN_rom = CN_rom,
    s_config = s_config, Tavg = sc$temps,
    n_years = n_years_sim, input_years = 1, input_month = 1
  )
})

summaries <- Map(function(sim) summarize_scenario(sim, Cin_total, H = H), sims)
```

## Figure 6 - CO2 emissions

```{r fig6}
fig6_df <- bind_rows(lapply(names(sims), function(nm) {
  sims[[nm]] %>%
    mutate(scenario = nm, t_yr = (step - 1) / 12) %>%
    filter(t_yr <= 10) %>%
    select(scenario, t_yr, em_CO2)
}))

ggplot(fig6_df, aes(t_yr, em_CO2, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Years after C input", y = "CO2-C emission (kg / month)", colour = NULL,
       title = "CO2 emission in the 10 years following the input of 1000 kg C") +
  theme(legend.position = "bottom")
```

As in the manuscript, the plant-material curves for C:N = 16:1 and C:N =
50:1 are identical (carbon turnover does not depend on the nitrogen content
of the input); the manure curve differs because a fraction of the manure C
enters the more slowly decomposing HUM pool immediately; and the cool-climate
curve is delayed and flattened relative to the standard climate.

## Figure 7 - N2O(-N) emissions

```{r fig7}
fig7_df <- bind_rows(lapply(names(sims), function(nm) {
  sims[[nm]] %>%
    mutate(scenario = nm, t_yr = (step - 1) / 12) %>%
    filter(t_yr <= 10) %>%
    select(scenario, t_yr, N2ON_emitted)
}))

ggplot(fig7_df, aes(t_yr, N2ON_emitted, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Years after C input", y = expression(N[2]*O*"-N emission (kg / month)"), colour = NULL,
       title = "N2O-N emission in the 10 years following the input") +
  theme(legend.position = "bottom")
```

Unlike CO2, the N2O curves now differ between C:N = 16:1 and C:N = 50:1
(more N enters with the C:N = 16:1 input), and the manure curve shows the
delayed second phase described in the text as N mineralises from the larger
HUM pool the manure feeds directly.

## Figure 8 - instantaneous warming from the sequestered-C CO2, vs. a pulse

```{r fig8}
years_out <- seq(0, H, by = 0.5)
pulse_CO2_mass <- Cin_total * gamma_C_CO2

fig8_df <- bind_rows(lapply(names(summaries), function(nm) {
  s <- summaries[[nm]]
  data.frame(scenario = nm, year = years_out,
             warming = instantaneous_warming_CO2(s$beta_t, years_out, total_co2_mass = pulse_CO2_mass))
}))
fig8_pulse <- data.frame(scenario = "Pulse (immediate) emission", year = years_out,
                          warming = instantaneous_warming_pulse(pulse_CO2_mass, years_out, gas = "CO2"))

ggplot(bind_rows(fig8_df, fig8_pulse), aes(year, warming, colour = scenario, linetype = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Years after input", y = expression("Instantaneous warming (W "*m^-2*")"),
       colour = NULL, linetype = NULL,
       title = "Instantaneous warming: sequestered-C CO2 vs. an immediate pulse") +
  theme(legend.position = "bottom") + guides(colour = guide_legend(nrow = 2))
```

Sequestration delays and spreads out the CO2 emission, so initial warming is
lower than a pulse of the same mass - but because so much of the input is
eventually re-emitted, the sequestered-C curves cross above the pulse curve
later on, exactly as described for Fig. 8 in the text.

## Figure 9 - instantaneous warming from the associated N2O, vs. a pulse

```{r fig9, fig.height=7, fig.width=9}
fig9_df <- bind_rows(lapply(names(summaries), function(nm) {
  s <- summaries[[nm]]
  data.frame(scenario = nm, year = years_out,
             warming = instantaneous_warming_N2O(s$eta_t, years_out, total_n2o_mass = s$total_N2O_kg))
}))
fig9_pulse <- bind_rows(lapply(names(summaries), function(nm) {
  s <- summaries[[nm]]
  data.frame(scenario = nm, year = years_out,
             warming = instantaneous_warming_pulse(s$total_N2O_kg, years_out, gas = "N2O"))
}))
fig9_df$series <- "Sequestered (delayed) N2O"
fig9_pulse$series <- "Pulse (immediate) N2O"

ggplot(bind_rows(fig9_df, fig9_pulse), aes(year, warming, colour = series)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~scenario, ncol = 2, scales = "free_y") +
  labs(x = "Years after input", y = expression("Instantaneous warming (W "*m^-2*")"), colour = NULL,
       title = "Instantaneous warming: delayed N2O vs. an immediate pulse of the same mass") +
  theme(legend.position = "bottom")
```

## Table 2 - net GWP effect over H = 100 years

```{r table2}
table2 <- bind_rows(lapply(names(summaries), function(nm) {
  s <- summaries[[nm]]
  data.frame(
    `Organic matter input` = nm,
    `C seq. offset (kg CO2e)` = round(s$C_offset_kgCO2e, 0),
    `N2O emission (kg CO2e)`  = round(s$N2O_effect_kgCO2e, 0),
    `Net effect (kg CO2e)`    = round(s$net_kgCO2e, 0),
    `Effective GWP of N2O`    = round(s$gwp_n2o_effective, 0),
    check.names = FALSE
  )
}))
kable(table2, caption = "Net GHG effect (kg CO2e) of a 1000 kg C input, H = 100 yr")
```

### Sanity checks against the manuscript text

```{r sanity-check}
paper_quoted <- data.frame(
  scenario = c("Plant (C:N 16:1)", "Plant (C:N 50:1)", "Manure", "Plant, cool climate"),
  paper_C_offset = c(-167, -167, -302, -283),
  paper_N2O_effect = c(252, 70, 238, 240),
  paper_eff_GWP = c(256, 222, 241, 244)
)
comparison <- table2 %>%
  rename(scenario = `Organic matter input`,
         model_C_offset = `C seq. offset (kg CO2e)`,
         model_N2O_effect = `N2O emission (kg CO2e)`,
         model_eff_GWP = `Effective GWP of N2O`) %>%
  left_join(paper_quoted, by = "scenario") %>%
  mutate(C_offset_pct_diff = 100 * (model_C_offset - paper_C_offset) / paper_C_offset,
         N2O_effect_pct_diff = 100 * (model_N2O_effect - paper_N2O_effect) / paper_N2O_effect)
kable(comparison[, c("scenario", "model_C_offset", "paper_C_offset", "C_offset_pct_diff",
                      "model_N2O_effect", "paper_N2O_effect", "N2O_effect_pct_diff")],
      digits = 1, caption = "Model output vs. numbers quoted in the manuscript text")
```

The residual few-percent gaps (mostly in the N2O effective GWP) are expected
given several pieces of this reconstruction had to be inferred from
prose rather than read directly off the manuscript's own equations (which
are embedded as un-OCR'able image objects in the .docx, not extractable
text) - see the header of `R/ghg_functions.R` for exactly which parts are
original C-TOOL/rCTOOL and which are this reconstruction.

## Bonus: repeated annual input vs. the conventional stock-change method

The Discussion section sketches, but does not fully specify, a scenario
comparing the conventional IPCC stock-change method against the AGWP
approach for a **repeated** annual C input (referred to as "Figure xx",
with no numeric detail given). The implementation below is a reasonable
reading of that paragraph, not a literal reproduction:

* **Conventional (stock-change) method**: each year's reported negative
  emission is $-(\Delta SOC_{stock}) \times 44/12$ - exactly the
  elemental-mass-balance approach the manuscript's Introduction critiques.
* **AGWP method**: the correct, time-invariant annual offset is the one
  already computed for a single 1000 kg C pulse ($GWP_{Cseq} \times C_{in}$),
  since (per the Discussion) "the GWP-weighted emissions ... will remain
  constant ... even if repeated each year".
* The **over-estimation ratio** is conventional / AGWP (ratio > 1 means the
  conventional method overstates the cooling effect that year).

```{r bonus}
repeat_years <- 60
sim_repeat <- run_ctool_pulse(
  Cin_total = Cin_total, is_manure = FALSE, f_man_humification = f_man_humification,
  CN_input = 16, CN_hum = CN_hum, CN_rom = CN_rom,
  s_config = s_config, Tavg = T_standard,
  n_years = repeat_years, input_years = 1:repeat_years, input_month = 1
)

soc_annual <- sim_repeat %>%
  group_by(year) %>%
  summarise(SOC_stock = last(FOM_top + HUM_top + ROM_top), .groups = "drop") %>%
  mutate(dSOC = SOC_stock - dplyr::lag(SOC_stock, default = 0),
         conventional_kgCO2e = -dSOC * gamma_C_CO2)

agwp_annual_offset <- summaries[["Plant (C:N 16:1)"]]$C_offset_kgCO2e

overestimate_df <- soc_annual %>%
  mutate(agwp_kgCO2e = agwp_annual_offset,
         overestimation_ratio = conventional_kgCO2e / agwp_kgCO2e)

ggplot(overestimate_df, aes(year, overestimation_ratio)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.8) +
  labs(x = "Year since repeated annual input began",
       y = "Over-estimation ratio\n(conventional / AGWP)",
       title = "Over-estimation of the conventional method for a repeated annual 1000 kg C input")

cat(sprintf("Year-1 over-estimation ratio: %.2f\n", overestimate_df$overestimation_ratio[1]))
```

## Session info

```{r session-info}
sessionInfo()
```
